---
title: "Homework 2"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
```

# Problem 1

```{r}
nyc_transit_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = "") |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(entry = case_match(
    entry,
    "YES" ~ TRUE,
    "NO" ~ FALSE
    )
  )
```

# Problem 2

```{r}
mr_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = 1,
    skip = 1,
    range = cell_cols("A:N")
  ) |> 
  janitor::clean_names() |> 
  filter(dumpster > 0) |> 
  mutate(
    homes_powered = (weight_tons*500)/30,
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "mr",
    year = as.numeric(year)
  )
```

```{r}
professor_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = 2,
    skip = 1,
    range = cell_cols("A:M")
  ) |> 
  janitor::clean_names() |> 
  filter(dumpster > 0) |> 
  mutate(
    homes_powered = (weight_tons*500)/30,
    trash_wheel = "professor",
    plastic_bottles = as.numeric(plastic_bottles)
  )
```

```{r}
gwynnda_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = 4,
    skip = 1,
    range = cell_cols("A:L"),
  ) |> 
  janitor::clean_names() |> 
  filter(dumpster > 0) |> 
  mutate(
    homes_powered = (weight_tons*500)/30,
    trash_wheel = "gwynnda"
  )
```

```{r}
trash_wheel_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df) |> 
  relocate(trash_wheel)
```

In the Trash Wheel dataset, there are `r nrow(trash_wheel_df)` observations across three trash wheels. Mr. Trash Wheel has `r nrow(mr_trash_df)` observations, Professor Trash Wheel has `r nrow(professor_trash_df)` observations, and Gwynnda the Good Wheel has `r nrow(gwynnda_df)` observations. Key variables in the Trash Wheel dataset include `trash_wheel`, `year`, `weight_tons`, `volume_cubic_yards`, and `homes_powered`. Data spans from `r min(trash_wheel_df$year)` to `r max(trash_wheel_df$year)`. On average, the trash wheels collected `r round(mean(trash_wheel_df$weight_tons), 2)` tons of trash and `r round(mean(trash_wheel_df$volume_cubic_yards), 2)` cubic yards of volume per dumpster, enough to power an estimated `r round(mean(trash_wheel_df$homes_powered))` homes for a day.

The total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_df$weight_tons)` tons.

The total number of cigarette butts collected by Gwynnda the Good Wheel in June of 2022 is `r sum(filter(gwynnda_df, year == 2022, month == "June")$cigarette_butts)` cigarette butts.

# Problem 3

```{r}
baseline_df =
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1,
    na = "."
  ) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  ) |> 
  rename(study_id = id) |> 
  filter(age_at_onset > current_age | is.na(age_at_onset))
```

When importing the baseline dataset, I skipped the first row and treated periods as missing values. Then, I recoded sex and APOE4 status to the appropriate labels, and renamed id to study_id so that the baseline dataset could be joined with the amyloid dataset. To meet the inclusion criteria of no MCI at baseline, I excluded participants whose onset age was at or before their baseline age, retaining those who developed MCI after baseline or remained MCI-free during follow-up.

After cleaning the dataset, there were `r nrow(baseline_df)` participants. Of these, `r sum(!is.na(baseline_df$age_at_onset))` (`r round(mean(!is.na(baseline_df$age_at_onset)) * 100, 2)`%) developed MCI. On average, the baseline age was `r round(mean(baseline_df$current_age))` years. Among women, the proportion of those who were APOE4 carriers was `r round(nrow(filter(baseline_df, sex == "female", apoe4 == "carrier")) / nrow(filter(baseline_df, sex == "female")), 2)`.

```{r}
amyloid_df =
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")
  ) |> 
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8,
    names_to = "year",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |> 
  mutate(
    year = replace(year, year == "baseline", "0"),
    year = as.numeric(year)
  )
```

When importing the amyloid dataset, I skipped the first row and treated Na and NA as missing values. I used the `pivot_longer()` function to reshape the data from wide to long format because visits were spread across multiple columns. Then, I recoded baseline to 0 and converted year to a numeric variable. 

```{r}
mci_df =
  left_join(baseline_df, amyloid_df, by = "study_id")
```

There were `r length(unique(mci_df$study_id))` participants who appeared in both the baseline and amyloid datasets. Key variables in the MCI dataset include `study_id`, `age_at_onset`, `year`, and `amyloid_ratio`. Among participants who developed MCI, the mean age at onset was `r round(mean(mci_df$age_at_onset, na.rm = TRUE))` years. The average amyloid ratio was `r round(mean(filter(mci_df, year == 0)$amyloid_ratio, na.rm = TRUE), 3)` at baseline, `r round(mean(filter(mci_df, year == 2)$amyloid_ratio, na.rm = TRUE), 3)` at year 2, `r round(mean(filter(mci_df, year == 4)$amyloid_ratio, na.rm = TRUE), 3)` at year 4, `r round(mean(filter(mci_df, year == 6)$amyloid_ratio, na.rm = TRUE), 3)` at year 6, and `r round(mean(filter(mci_df, year == 8)$amyloid_ratio, na.rm = TRUE), 3)` at year 8.

```{r}
write_csv(mci_df, "data/mci_df.csv")
```